Put Your Money Where Your Mouth Is: Prediction Markets and Climate Change

### 2016-03-19 paper.Rmd
small paper work

---
title: 'Put Your Money Where Your Mouth Is: Prediction Markets and Climate Change'
author: "John J. Nay, Martin Van der Linden, Jonathan Gilligan"
date: "March 17, 2016"
output:
  pdf_document:
    keep_tex: yes
  html_document: default
bibliography: library.bib
---
```{r fig_ref, echo=FALSE}
source("utils/fig_ref.R") # tabRef and figRef functions
source("utils/multi_plot.R")
``` 
```{r, caption_setting, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(knitr)
knit_hooks$set(htmlcap = function(before, options, envir) {
  if(!before) {
    paste('<p class="caption">',options$htmlcap,"</p>",sep="")
    }
    })
# small margins
knit_hooks$set(small_mar = function(before, options, envir) {
    if (before) par(mar = rep(0.0001, 4))  # smaller margins all around
})
# opts_chunk$set(dev="tiff", 
#                dev.args=list(compression="lzw", type="cairo"),
#                dpi=300)
```
```{r multi_plot, echo=FALSE}
source("utils/multi_plot.R")
library(ggplot2)
theme_set(theme_classic())
library(pander)
```

@@

The climate change debate has become strongly polarized over the past two decades. Although the scientific consensus on the anthropogenic nature of climate change strongly increased, beliefs about climate change did not evolve much within the public [@Vandenbergh2013f]. In addition, the divide on anthropogenic climate change between liberals and conservatives has grown steadily as the question is becoming increasingly politicized and potentially disconnected from scientific evidence. The costs of misinformed climate policies are high. If anthropogenic climate change is a myth but is taken to be true, public resources will be spent on unnecessary efforts. On the other hand, if climate change is human-induced but not recognized as so, the costs of inaction could be devastating. Given that effective climate policies require long-lasting measures to be taken, the importance of an accurate consensus on the issue is manifest.

Attempts to foster such consensus face many social and psychological challenges, some of which could be adressed by creating climate prediction markets where participants can "put their money where their mouths are" [@Hsu2011; @Vandenbergh2013f]. The idea of using prediction markets to efficiently aggregate information about uncertain events has been widely discussed [@Horn2014c]. Prediction markets have been shown to have interesting theoretical properties [@Set1998; @Hanson2012] and to perform well in terms of prediction accuracy and information aggregation in experiments [@Hanson2006; @Healy2010], simulation models [@Klingert2012c; @Jumadinova2011a], and the real-world [@Wolfers2006]. We simulate a prediction market where traders exchange securities related to climate outcomes to explore whether, and under what market and climate conditions, prediction markets may be useful for increasing convergence of climate beliefs. The ultimate goal is to initiate a computational design process for effective climate prediction markets.

@@

@Tseng2010a create an agent-based model (ABM) of a continuous double auction market with various types of market strategies, including two variants of the zero-intelligence agent. They compare the behavior of their simulation with data from a prediction market for the outcome of political elections. They find that despite their simplicity, zero-intelligence agents capture some salient features of real market data. @Klingert2012c reach similar conclusions with respect to experimental data from @Hanson2006. In their simulation, @Klingert2012c compare continuous double auctions with logarithmic market scoring rules and find that the latter generally outperform continuous double auctions in terms of prediction accuracy. In some variants of the two last models, agents' strategies feature learning, but these are rather basic and unrelated to a formal predictive model.

@Jumadinova2011a conducted research more similar to ours, as they explicitly include information about uncertain events in their model. @Jumadinova2011a study a continuous double auction model in which agents update their beliefs about the uncertain events based on information on the event. New beliefs are modeled as a weighted sum of the last period's beliefs and market prices. The agents' information set changes every period. The better the information set, the more likely the agent is to put higher weight on last period's prices when revising their beliefs. Little structure is imposed on the information set and the way it is used to generate the weights in the update function. Therefore, the model does not allow for investigating the convergence of predictive models. @Ontanon2009 study the effect of deliberation on a simulated prediction market. They assume that agents use Case-Based Reasoning [@Aamodt1994] to reason and debate about uncertain outcomes with their neighbors in a social network. Although the model could in theory be used to assess convergence of predictive models by looking at the arguments agents adopt after the deliberation, this is not done by the authors. Also, the way beliefs are formed and revised through argumentation is too abstract for studying climate prediction markets, where beliefs should be based on statistical time series models.

@@

From a public policy perspective, changing the explanatory models of market participants is one of the most important roles that prediction markets might play. This is perhaps the most important social benefit of prediction markets given the comparable predictive power of statistical supervised learning models, which can provide information at a much lower cost than creating and maintaining a prediction market [@goel_prediction_2010]. Effective climate policy not only requires an accurate consensus on future climate *outcomes*. It also requires an accurate consensus on the causal *mechanisms* influencing such outcomes. If people agree that temperature will rise, but some believe it will be due to greenhouse gases, while others believe that it will be caused by increased solar activity, inconsistent and ineffective policies may be implemented.

@@

The sensitivity analysis averages over time (and model replicates) and thus masks time trends. We randomly drew from the above distributions for ideology, risk tolerance, and the number of traders, and set conducted an experiment crossing 95% and 5% percentile values for the number of edges per trader and the segmentation of the social network and both values for the true model. We then collected the time series of convergence across these eight designs (`r figRef("time")`).

@@

Our code for the model is available at [johnjnay.com/predictMarket](http://johnjnay.com/predictMarket). This project is a computational test-bed for public policy design.

>>>>

---
title: 'Put Your Money Where Your Mouth Is: Prediction Markets and Climate Change'
author: "John J. Nay, Martin Van der Linden, Jonathan Gilligan"
date: "March 19, 2016"
output:
  pdf_document:
    keep_tex: yes
  html_document: default
bibliography: library.bib
---
```{r fig_ref, echo=FALSE}
source("utils/fig_ref.R") # tabRef and figRef functions
source("utils/multi_plot.R")
``` 
```{r, caption_setting, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(knitr)
knit_hooks$set(htmlcap = function(before, options, envir) {
  if(!before) {
    paste('<p class="caption">',options$htmlcap,"</p>",sep="")
    }
    })
# small margins
knit_hooks$set(small_mar = function(before, options, envir) {
    if (before) par(mar = rep(0.0001, 4))  # smaller margins all around
})
# opts_chunk$set(dev="tiff", 
#                dev.args=list(compression="lzw", type="cairo"),
#                dpi=300)
```
```{r multi_plot, echo=FALSE}
source("utils/multi_plot.R")
library(ggplot2)
theme_set(theme_classic())
library(pander)
```

@@

The climate change debate has become strongly polarized over the past two decades. Although the scientific consensus on the anthropogenic nature of climate change strongly increased, beliefs about climate change did not evolve much within the public [@Vandenbergh2013f]. In addition, the divide on anthropogenic climate change between liberals and conservatives has grown steadily as the question is becoming increasingly politicized and potentially disconnected from scientific evidence. The costs of misinformed climate policies are high. If climate change is not human-induced but is believed to be so, public resources will be spent on unnecessary efforts. On the other hand, if climate change is human-induced but not recognized as so, the costs of inaction could be devastating. Given that effective climate policies require long-lasting measures to be taken, the importance of an accurate consensus on the issue is manifest.

Attempts to foster such consensus face many social and psychological challenges, some of which could be adressed by creating climate prediction markets where participants can "put their money where their mouths are" [@Hsu2011; @Vandenbergh2013f]. The idea of using prediction markets to efficiently aggregate information about uncertain event outcomes has been widely discussed [@Horn2014c]. Prediction markets have interesting theoretical properties [@Set1998; @Hanson2012], and perform well in terms of prediction accuracy and information aggregation in experiments [@Hanson2006; @Healy2010] simulation models [@Klingert2012c; @Jumadinova2011a] and the real-world [@Wolfers2006]. We simulate a prediction market where traders exchange securities related to climate outcomes to explore whether, and under what market and climate conditions, prediction markets may be useful for increasing convergence of climate beliefs. Our work can be extended as part of a computational design process for effective climate prediction markets.

@@

@Tseng2010a create an agent-based model (ABM) of a continuous double auction market with multiple market strategies, including two variants of the zero-intelligence agent. They compare the behavior of their simulation with data from a prediction market for the outcome of political elections. They find that despite their simplicity, zero-intelligence agents capture some salient features of real market data. @Klingert2012c reach similar conclusions with respect to experimental data from @Hanson2006. In their simulation, @Klingert2012c compare continuous double auctions with logarithmic market scoring rules and find that the latter generally outperform continuous double auctions in terms of prediction accuracy. In some variants of the two last models, agents' strategies feature learning, but these are rather basic and unrelated to a formal predictive model.

@Jumadinova2011a conducted research similar to ours, as they explicitly include information about uncertain events in their model. @Jumadinova2011a study a continuous double auction model in which agents update their beliefs about uncertain events based on information on the event. New beliefs are modeled as a weighted sum of the last period's beliefs and market prices. The agents' information set changes every period. The better the information set, the more likely the agent is to put higher weight on last period's prices when revising their beliefs. Little structure is imposed on the information set and the way it is used to generate the weights in the update function. Therefore, the model does not allow for investigating the convergence of predictive models. @Ontanon2009 study the effect of deliberation on a simulated prediction market. They assume that agents use Case-Based Reasoning [@Aamodt1994] to debate uncertain outcomes with their neighbors in a social network. Although the model could in theory be used to assess convergence of predictive models by looking at the arguments agents adopt after the deliberation, this is not done by the authors. Also, the way beliefs are formed and revised through argumentation is too abstract for studying climate prediction markets, where beliefs should be based on statistical time series models.

@@

From a public policy perspective, changing the explanatory models of market participants is one of the most important roles that prediction markets might play. This is perhaps the most important social benefit of prediction markets given the comparable predictive power of statistical supervised learning models, which can provide information at a much lower cost than creating and maintaining a market [@goel_prediction_2010]. Effective climate policy not only requires an accurate consensus on future climate *outcomes*. It also requires an accurate consensus on the causal *mechanisms* influencing such outcomes. If people agree that temperature will rise, but some believe it will be due to greenhouse gases, while others believe that it will be caused by increased solar activity, inconsistent and ineffective policies may be implemented.

@@

The sensitivity analysis averages over time (and model replicates) and thus masks time trends. We randomly drew from the above distributions for ideology, risk tolerance, and the number of traders, and set conducted an experiment crossing 95% and 5% percentile values for the number of edges per trader and the segmentation of the social network and both values for the true model. We then collected the time series of convergence across these eight designs for visualization of the distributions of convergence over time (`r figRef("time")`).

@@

All code for the model is available at [johnjnay.com/predictMarket](http://johnjnay.com/predictMarket). This project is a computational test-bed for public policy design.
